What We Eval in the Shadows

### 2021-06-29 paper/main.tex
Update 2.3 section

R exposes a rich reflective interface with functions, \c{eval},
\c{evalq}, \c{eval.parent}, and \c{local}. The following simplifies details not
relevant to this paper, the interested reader should consult \citet{hadley}.

\begin{lstlisting}
 eval <- function(e, envir = parent.frame(),
                   encl = if(is.list(envir)) parent.frame() else baseenv()) ...
\end{lstlisting}
The most general is \eval. Parameter \c e is the \c{expression} to evaluate,
\c{envir} is the evaluation environment, and \c{encl} is used to look up
variables not found in \c{envir}. The default value of the latter is either the
caller's environment or the base environment \Eval can take several types for \c e, for
our purposes on the \c{expression}. These can be thought of as the abstract
syntax trees returned by the parser. An \c{expression} can be obtained by
calling \c{quote} or \c{parse} with a string:
\begin{lstlisting}
 > quote(a + b)
 # a + b
 > parse(text="a+b")
 # expression(a+b)
\end{lstlisting}
Usually, expressions are obtained from promises using \c{substitute}. Consider a
function called with \c{x = a+b}, then \c f returns an expression object.
\begin{lstlisting}
 > f <- function(x) substitute(x)
 > f(a+b)
 # a+b
 > substitute(a+b,list(a=1,b=2))
 # 1+2
\end{lstlisting}
The call to \c{substitute(x)} extracts the unevaluated expression from the
promise \c x. Function \c{substitute(e,envir)} takes two arguments, the second
is used to substitute free variables in \c e.
\begin{lstlisting}
 > environment()
 # <environment: R_GlobalEnv>
\end{lstlisting}
The \c{environment} function returns the current environment. Environments nest,
each has a parent. When creating a new environment with \c{new.env}, the parent
is the current environment. Environment chains can be traversed with
\c{parent.env}, until \c{emptyenv} is reached. The top-level environment is
\c{.GlobalEnv}, it has parents that represent the packages that have been
loaded. One can also directly read, modify or create new bindings, given any
environment:
\begin{itemize}
\item \c{envir\$v} and \c{get("v",envir=envir)}: read  \c{v} from \c{envir};
\item \c{envir\$v<-2} and \c{assign("v",2,envir=envir)}: store 2 in \c{v}, if
  not found, \c{v} is created in \c{envir}.
\end{itemize}
\noindent Environments are used as hash maps as they have reference semantics
and a built-in string lookup. Functions \c{parent.frame} and \c{sys.frame}
return environments further up the call stack.
 \begin{lstlisting}
  evalq <- function(e,envir,encl) eval(quote(e),envir,encl)
  eval.parent <- function(e,n) eval(e,parent.frame(n))
  local <- function(e,envir,encl) evalq(e,new.env())
\end{lstlisting}
The three variants can be expressed with \eval: \c{evalq} quotes its
argument to prevent evaluation in the current environment,
\c{eval.parent(e,n)} evaluates \c{e} in the environment of the {\tt n}-th
caller of this function; finally, \c{local} evaluates \c{e} in a new environment
to avoid polluting the current one.

@@

The dynamic analysis is performed by \rdyntrace, a modified R virtual machine
based on GNU R 4.0.2 that exposes low-level callbacks for a variety of runtime
events~\cite{oopsla19b}. The tracer registers callbacks to all \eval variants
and a few additional functions to assist in locating the origin of \eval
arguments. For example, we taint the results of calls to \c{parse} and
\c{match.call} (reflects on current call). We also capture calls to R API for
dynamic code loading. Next, we subscribe to the events related to a variable
definition and assignment, allowing us to record side effects that happen in
environments while evaluating code in \eval. One challenge was that R only
provides source references for block surrounded by braces. Thus, a function
whose body is \c{eval(x)} would no usable debug information, while the same
expression surrounded by braces would be fine. This is unfortunately not easily
fixed. We extended the dynamic analysis tool to attach synthetic source code
references to all \eval call sites by traversing ASTs. The tracer is implemented
as an R package in 3.2K C++ and 1.3K of R code. For performance reasons, most of
the tracing is done in C++. While in theory, the implementation is relatively
straightforward, not so in practice:
%
\begin{compactitem}[---]

>>>>

The \eval function in R takes three parameters: an \c{expression} to evaluate
(\c{e}), an environment where to evaluate (\c{envir}), and an enclosure
(\c{encl}) that is used to look up objects not found in \c{envir}.\footnote{Next to the \eval function R
  offers three other variants:
  \begin{inparaenum}[1.]
    \item \c{evalq} automatically quotes passed expression---a shorthand to
      \c{eval(quote(...))},
    \item \c{eval.parent(e, n)} specifies the evaluating
      environment in terms of number of call frames (\c{n}) to go back---a shorthand
      to \c{eval(e, parent.frame(n))}, and
    \item \c{local} evaluates \c{e} in a fresh
      environment---a shorthand to \c{evalq(e, new.env())}.
  \end{inparaenum} }
%
\begin{lstlisting}
 eval <- function(e, envir = parent.frame(),
                   encl = if(is.list(envir)) parent.frame() else baseenv()) ...
\end{lstlisting}

\noindent The expression passed to \eval can be thought of as an abstract
syntax tree. It can be parsed from a string, created manually or using
reflection:

\vspace*{-2mm}
\begin{figure}[htb]
\begin{minipage}{.49\textwidth}
\begin{lstlisting}[caption={Examples of calls producing expression \c{a+b}}]
  parse(text="a+b")
  quote(a+b)
  call("+", quote(a), quote(b))

  # reflect promise
  f <- function(x) substitute(x)
  f(a+b)
  \end{lstlisting}
\end{minipage}
\begin{minipage}{.49\textwidth}
  \begin{lstlisting}[caption={Example of a call reflection}]
  f <- function(x, y) {
    mc <- match.call() # reflect curr. call
    mc[[1]] <- as.name("g")
    mc[["x"]] <- 2
    mc
  }
  f(1,2) # returns a call g(2,y)
  \end{lstlisting}
\end{minipage}
\end{figure}

\vspace*{-2mm}
\noindent The call to \c{substitute(x)} extracts the unevaluated expression from the
promise \c{x}. The function \c{match.call} returns an expression representing the current call.

R is generous when it comes to what is allowed to be passed as environment to
\eval. Besides ordinary environments, it also accepts a list or a data
frame (using element names or column names for variable resolution) or an
integer \c{n} (in which case it would use the \c{n.} call frame). If no
environment is specified, then the default is the environment where the call to
\eval was made (as returned by \c{parent.frame}).
%
Environments nest, each has a parent. When creating a new environment with
\c{new.env}, the parent is the current environment. Environment chains can be
traversed with \c{parent.env}, until \c{emptyenv} is reached. The top-level
environment is \c{.GlobalEnv}, it has parents that represent the packages that
have been loaded. One can also directly read, modify or create new bindings,
given any environment:
%
\begin{lstlisting}
  # reading                                # writing
  envir$v                                  envir$v <- 2
  get("v", envir=envir)                    assign("v", 2, envir=envir)
\end{lstlisting}
%
\noindent Environments are used as hash maps as they have reference semantics
and a built-in string lookup.

@@

The dynamic analysis is performed by \rdyntrace, a modified R virtual machine
based on GNU R 4.0.2 that exposes low-level callbacks for a variety of runtime
events~\cite{oopsla19b}. The tracer registers callbacks to all \eval variants
and a few additional functions to assist in locating the origin of \eval
arguments. For example, we taint the results of calls to \c{parse} and
\c{match.call}. We also capture calls to R API for dynamic code loading. Next,
we subscribe to the events related to a variable definition and assignment,
allowing us to record side effects that happen in environments while evaluating
code in \eval. One challenge was that R only provides source references for
block surrounded by braces. Thus, a function whose body is \c{eval(x)} would no
usable debug information, while the same expression surrounded by braces would
be fine. This is unfortunately not easily fixed. We extended the dynamic
analysis tool to attach synthetic source code references to all \eval call
sites by traversing ASTs. The tracer is implemented as an R package in 3.2K C++
and 1.3K of R code. For performance reasons, most of the tracing is done in
C++. While in theory, the implementation is relatively straightforward, not so
in practice:
%
\begin{compactitem}[---]
